```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
Stats of NBA Players by Hyun Jin Lee
========================================================
#Abstract

NBA is one of the premier sports organization, with lots of young aspiring
athletes trying to make it to the "Biggest Stage of All". So, what traits make
a great athelete? Basketball once was deemed a sport of bigs, such as Shaq, who
dominated the paint zone. Now, the era is defined by a term, "small ball" - 
smaller and agile guards, with skills. What makes an athlete a great scorer? 
We are here to find out.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(dplyr)
library(ggplot2)
library(gridExtra)
library(RColorBrewer)
library(memisc)
```


#Dataset

The dataset is from Kaggle, with the work done by DrGuillermo scraping
Basketball-reference[https://www.basketball-reference.com]. His data includes
all players from 1950 to 2016, but as we are only looking at the modern NBA
atheletes, and 2017 season is incomplete in this dataset, our analysis will 
revolve around the 2016-17 season. It includes all players that were on the 
roster during 2016-17 season and played more than a minute throughout the season.

```{r echo=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
pl = read.csv("Players.csv")
st = read.csv("Seasons_Stats.csv")
df = left_join(st, pl, by="Player")
df <- df[!is.na(df$Player),]
df <- subset(df, Year == 2016)
#df <- subset(df, Year >= 2000)
df$PPM <- df$PTS / df$MP
df$APM <- df$AST / df$MP
df$SPM <- df$STL / df$MP
df$ORPM <- df$ORB / df$MP
df$TPM <- df$TOV / df$MP
df$FPM <- df$PF / df$MP
drops <- c("Year", "blanl", "blank2", "collage", "birth_city", "born", "Tm", "birth_state")
df <- df[, !(names(df) %in% drops)]
df = subset(df, MP > 0)
```



# Univariate Plots Section

```{r echo=FALSE, Dimension}
dim(df)
```

```{r echo=FALSE, Columns}
str(df)
```

```{r echo=FALSE, Summary}
summary(df)
```

Our dataset consists of 56 variables, with the majority of the metrics coming
from https://www.basketball-reference.com/about/glossary.html.

```{r echo=FALSE, Multi_Position_Players}
subset(df, Pos == "PF-C" | Pos == "SG-SF")
which(df$Pos == "PF-C")
#Channing Frye
df[166, "Pos"] <- "PF"

#extracting Weems
df = subset(df, Pos != "SG-SF")
```

There are only two players, out of 578, who have a dual-position as their
position. For the consistency, these outliers were addressed manually. As Frye 
is registered as forward on official NBA website, his position was adjusted as
such. However, Sonny Weems is a different case - he was, indeed, a tweener, and
was assigned both SG and SF for his role. Since he is an extreme outlier, and
the sample size is big enough that the exemption of one sample will not change
the statistics dramatically, I've decided exclude Sonny Weems sample from the 
dataset.

```{r echo=FALSE, Games}
ggplot(aes(x=G), data = df) +
  geom_histogram(binwidth = 4) +
  xlab("Games")
```

```{r echo=FALSE, PTS}
ggplot(aes(x=PTS), data = df) +
  geom_histogram(binwidth = 50) +
  xlab("Total Points")
```

```{r echo=FALSE, PTS_Summary}
summary(df$PTS)
```

```{r echo=FALSE, Log_PTS}
ggplot(aes(x=PTS+1), data = df) +
  geom_histogram(bins=200) +
  scale_x_log10() +
  xlab("Total Points")
```

Transformed the long-tail data using log10, to better understand the
distribution of total points between players. There seems to be a unimodal peak
near 1000 region. Generally, more minutes played, more chance for points an 
athelete gets. So, how about the minutes played?

```{r echo=FALSE, MP}
ggplot(aes(x=MP), data = df) +
  geom_histogram(binwidth = 50) +
  xlab("Minutes Played")
```

Both total points and minutes played show a similar pyramidic pattern, with only
a handful making their way up top. Then, I wonder, would points per minute 
played be drastically different from those patterns?

```{r echo=FALSE, PPM}
ggplot(aes(x=PPM), data = df) +
  geom_histogram(binwidth = .05) +
  xlab("Points Per Minute")
```

Quite surprisingly, points per minute actually shows a unimodal, normal-like 
curve with the mean around 0.39, with a few glaring outliers.

```{r echo=FALSE, PPM_Summary}
summary(df$PPM)
```

Now that we examined PTS & minutes played, turning it into point per minute, 
I want to examine other indicators that conjure up the offense. For field goals,
I believe that the FG% is a better indicator of a scorer's ability than numbers 
made or attempted, so I will examine that as a main feature of field goal.

```{r echo=FALSE, FG}
ggplot(aes(x=FG.), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("Field Goal %")
```


```{r echo=FALSE, FG_Summary}
summary(df$FG.)
```

```{r echo=FALSE, NO_FG}
subset(df, is.na(df$FG.))
```

The NAs in FG. indicate that the athlete did not take any field goals at 
all. They have 0 attempts, and 0 made.

Similarly, free throw % and 2-pointer / 3-pointer % will be examined instead of
made and attempted.

```{r echo=FALSE, FT}
ggplot(aes(x=FT.), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("Free Throw %")
```

```{r echo=FALSE, FT_Summary}
summary(df$FT.)
```

```{r echo=FALSE, X2PT}
ggplot(aes(x=X2P.), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("2-Point %")
```

```{r echo=FALSE, X2PT_Summary}
summary(df$X2P.)
```

```{r echo=FALSE, X3PT}
ggplot(aes(x=X3P.), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("3-Point %")
```

```{r echo=FALSE, X3PT_Summary}
summary(df$X3P.)
```

All four of the field goal related features show a normal-like curve, with some
outliers. However, the number of players with 3-pointer% at 0 are unusually 
high.

```{r echo=FALSE, No_X3PT}
count(subset(df, df$X3P. == 0))
```
```{r echo=FALSE, X3PTA_Summary}
summary(subset(df, df$X3P. == 0)[,c("X3P", "X3PA", "X3P.")])
```

There are 121 athletes with 3-pointer % of 0%. The number of attempts for these
athletes are generally very small, and all ended up with 0 made.

Since we checked the field goals already, let's check other numbers that conjure
an offense - assist, steal, offensive rebound, and turnover.

```{r echo=FALSE, APM}
ggplot(aes(x=APM), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("Assists per Minute")
```

```{r echo=FALSE, APM_Summary}
summary(df$APM)
```

```{r echo=FALSE, SPM}
ggplot(aes(x=SPM), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("steals per minute")
```

```{r echo=FALSE, SPM_Summary}
summary(df$SPM)
```


```{r echo=FALSE, ORPM}
ggplot(aes(x=ORPM), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("offensive rebound per minute")
```

While assist has formed a normal-like distribution, both steal and offensive
rebound forms a long-tailed graph. Lastly, we will examine the profiles of the
athletes, with height, weight, age, and position.

```{r echo=FALSE, ORPM_Summary}
summary(df$ORPM)
```


```{r echo=FALSE, TPM}
ggplot(aes(x=TPM), data = df) +
  geom_histogram(binwidth = 0.01) +
  xlab("turnover per minute")
```

```{r echo=FALSE, TPM_Summary}
summary(df$TPM)
```


```{r echo=FALSE, height}
ggplot(aes(x=height), data = df) +
  geom_histogram(binwidth = 5)
```

```{r echo=FALSE, weight}
ggplot(aes(x=weight), data = df) +
  geom_histogram(binwidth = 5)
```

```{r echo=FALSE, height_weight}
ggplot(aes(x=height / weight), data = df) +
  geom_histogram(binwidth = 0.05)
```

height-weight ratio has a normal-like distribution.

```{r echo=FALSE, Age}
ggplot(aes(x=Age), data = df) +
  geom_histogram(binwidth = 1)
```

```{r echo=FALSE, Position}
ggplot(aes(x=Pos), data = df) +
  geom_bar()
```

Lack of multi-position players indicates that in modern basketball, position is
very specific and well-defined.



# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

There are 578 players in the dataset with 56 features.
Position is a factor variable with the following level.
```{r echo=FALSE, Pos_level}
levels(df$Pos)
```
 "C", "C-F", "C-PF", "C-SF", "F", "F-C", "F-G",  "G",  "G-F", "PF", "PF-C",
"PF-SF","PG", "PG-SF", "PG-SG", "SF", "SF-PF", "SF-PG", "SF-SG", "SG", "SG-PF", 
"SG-PG" "SG-SF" 

### What is/are the main feature(s) of interest in your dataset?

The main features in the data set are points per minute and field goal 
percentage, which is divisible into two-pointer % and three-pointer %.
I want to determine which features are the best for predicting the points per
minute for a player. I suspect that the field goal percentage, along with some 
other variables, will affect the points per minute, thereby help build a 
predictive model for the points scored.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I believe that the performance features that I've explored above, such as 
steals, and offensive rebounds, will support my investigation into points per
minute and field goal percentage.

### Did you create any new variables from existing variables in the dataset?
Points per minute(PPM) variable was derived from (total points scored / minutes
played). This is a better indicator of the athlete's proficiency in scoring than
total points scored, which can be biased by the length of their playing time.
Also, assist / steal / offensive rebounds / turnover are all converted to a per
minute metric.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

Before we dive into the variables, let's make sense of the correlation of our 
main variable, points per minute(PPM), to other possibly affecting variables.


```{r echo=FALSE, Bivariate_Plots}
M = cor(df[, c("PPM", "Age", "FG.", "FT.", "X3P.", "X2P.", "ORPM", "APM", "SPM", "TPM", "FPM", "height", "weight")], use = "pairwise.complete.obs")
M[, "PPM"]
```

Pearson's correlation of variables against PPM shows that there two indicators 
showing modest correlation - field goal percentage (FG.) and turnovers per
minute(TPM). 

There are three interesting things to note: 
1. Field goal percentage as a whole has much higher correlation, when compared 
to the broken down components of field goal percentage : 2-pointer percentage 
and 3-pointer percentage.

2. Turnover has a relatively high correlation with points per minute, which is
quite surprising - generally, one would think that having more turnovers would
undermine one's ability to score.

3. Physical profile such as height & weight, is seemingly unrelated with the
athletes' point scoring ability. 

Let's examine the first point, with field goal percentage, 2-pointer percentage, 
 and 3-pointer percentage in relation to point per minute.

```{r echo=FALSE, FG_PPM}
ggplot(aes(x=FG., y=PPM), data = df) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm") +
  scale_y_continuous(limits = c(0, 1)) +
  xlab("Field Goal Percentage")
```

To take a closer look at the main area, let's zoom in:

```{r echo=FALSE, FG_PPM_Zoom}
ggplot(aes(x=FG., y=PPM), data = df) +
  geom_jitter(alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0.25, 0.65)) +
  scale_y_continuous(limits = c(0, 1)) + 
  xlab("Field Goal Percentage")
```

```{r echo=FALSE, X2P_PPM}
ggplot(aes(x=X2P., y=PPM), data = df) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm") +
  xlab("2-Pointer Percentage")
```


```{r echo=FALSE, X3P_PPM}
ggplot(aes(x=X3P., y=PPM), data = df) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm") + 
  xlab("3-Pointer Percentage")
```

Field Goal Percentage is defined by (X2P + X3P) / (X2PA + X3PA), thus is related
directly to 2-pointer percentage, defined by (X2P / X2PA), and 3-pointer 
percentage, defined by (X3P / X3PA). Thus, when linear modelling, including 
field goal percentage along with 2-pointer percentage and 3-pointer percentage
will likely cause a colinearity issue. Since Field Goal Percentage has higher
correlation with point per minute, it might as well be one of our choice of the
variables in linear model.

Now, we can address our second point in question - that is, the question of 
high correlation between turnover per minute(TPM) and point per minute. It is
quite counter-intuitive, as turnover generally means that you coughed up your
ball, and in general, ball handling is emphasized in the pro scene.

```{r echo=FALSE, TPM_PPM}
ggplot(aes(x=TPM, y=PPM), data = df) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm") + 
  xlab("Turnover per Minute")
```

A positive tendency is quite strong. Let us examine the concentrated area, from 
around 0.01 to 0.1.

```{r echo=FALSE, TPM_PPM_Zoom}
ggplot(aes(x=TPM, y=PPM), data = df) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm") + 
  scale_x_continuous(limits = c(0.01, 0.1)) +
  xlab("Turnover per Minute")
```

The tendency seems even clearer, even though there seems to be a spread towards
the higher end, a bit like a cone shape. Let's examine some of the high TOM 
personnels, in top 90% percentile.

```{r echo=FALSE, High_TPM}
High_T = arrange(subset(df, TPM > quantile(df$TPM, .9, na.rm = TRUE)), desc(PPM))
head(High_T, 10)
```

Except for some the outliers whose game stinct was quite brief, you find some 
interesting names - Curry, KD, DeMarcus, Harden, LeBron James, and so on. These
athletes are entrusted with leading the charge, the primary ball handlers. Thus,
high turnover per minute is quite natural to them.

This is the summary of point per minute, for athletes with high turnovers per 
minute:

```{r echo=FALSE, PPM_High_TPM_Summary}
summary(High_T$PPM)
```

This is the summary of point per minute, over whole dataset:

```{r echo=FALSE, PPM_Summary2}
summary(df$PPM)
```

The difference is quite staggering.

```{r echo=FALSE, Cor_TPM}
M = cor(df[, c("PPM", "Age", "FG.", "FT.", "X3P.", "X2P.", "ORPM", "APM", "SPM", "TPM", "FPM", "height", "weight")], use = "pairwise.complete.obs")
M[, "TPM"]
```

Furthermore, the correlation between turnover per minute and assist per minute 
is even higher than that between turnover per minute and points per minute. This
is also quite reasonable, as our list of top turnover per minute athletes were
the primary ball handlers for the team, leading attack and responsible for 
scoring and creating for teammates. 

How about the positions? We know that positions define the role, with the 
centers and power forwards operating nearer to the goal. Guards, generally,
operate on the perimeter. Does the position of an athlete have any impact?

```{r echo=FALSE, Density_FG_Pos}
ggplot(aes(x=FG.), data = df) +
  geom_density(aes(color = Pos)) +
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Positions', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

```{r echo=FALSE, Density_PPM_Pos}
ggplot(aes(x=PPM), data = df) +
  geom_density(aes(color = Pos)) +
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Positions', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

This density plot shows an odd trend in field goal percentage and point per
minute concerning centers. The centers, while having point per minute density 
similar to other positions, have higher field goal percentage compared to
the other positions.

```{r echo=FALSE, BOX_FG_Pos}
ggplot(aes(x=Pos,y=FG.), data = df) +
  geom_boxplot() +
  ylab("Field Goal Percentage")+
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Positions', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

This difference is even more stark when looking at box plot for field goal
percentage, segmented by positions. Centers have a very high inter-quartile
range, with the median way above the other positions. This relationship 
between field goal percentage, point per minute, and position poses an 
interesting question, and thus will be examined further in the multivariate 
section.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

My main feature of interest, points per minute, had moderate positive 
correlation with field goal percentage (FG.) and turnovers per minute(TPM),
both positive. Also, it had positive correlation with free throw percentage, as
well.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Turnovers per minute(TPM) has a strong positive correlation with assist per 
minute(APM), which is quite plausible as the athletes whose turnovers per minute
was the highest were the primary ball handlers for their team. Also, centers 
had higher field goal percentage relatively.

### What was the strongest relationship you found?

The strongest relationship that I found was that between field goal percentage
and points per minute. The correlation between field goal percentage and points
per minute was higher than the broken down domains of 3-pointer percentage and 
2-pointer percentage. This is quite an interesting phenomenon.


# Multivariate Plots Section


```{r echo=FALSE, PPM_FG_POS}
ggplot(aes(x=FG., y=PPM), data = df) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm")+
  facet_wrap(~Pos) + 
  xlab("Field Goal Percentage")
```

As you can see, for most of the positions, field goal percentage seems to have
a positive linear correlation with points per minute. However, for center 
position, the field goal percentage does not show such a correlation. The 
correlation between points per minute and field goal percentage, for centers, is
small as shown below:

```{r echo=FALSE, C_PPM_FG_cor}
C = subset(df, Pos == "C")
cor(C$PPM, C$FG., use = "complete.obs")
```

Excluding the centers, the correlation between points per minute and field goal 
percentage becomes quite strong:

```{r echo=FALSE, D_PPM_FG_cor}
D = subset(df, Pos != "C")
cor(D$PPM, D$FG., use = "complete.obs")
```

This is greater than the correlation as whole.

This, along with the plots of field goal percentage and positions done at the 
bivariate section, implies that centers generally have higher field goal 
percentage when compared to other positions. It is reasonable, as centers, 
generally, shoot a lot closer to the rim than most of the other positions.

```{r echo=FALSE, PPM_FG_Pos_Color}
ggplot(aes(x=FG., y=PPM), data = df) +
  geom_point(aes(color = Pos)) +
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Positions', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

When visualized, centers reside much closer to the lower-right cluster of 
field goal percentage - point per minute plot, with higher field goal percentage
and lower point per minute. Also, higher spread compared to other positions, if 
not regarding the outliers.

The field goal percentage, as a whole, shows a weak linear relationship with the 
points per minute, meaning that it can be, indeed, a variable which can be used
to predict the point per game. However, given the big spread, the statistical
power of the model will be quite weak.

```{r echo=FALSE, LM}
m1 <- lm((PPM ~ FG.), data = df)
m2 <- update(m1, ~ . + TPM)
m3 <- update(m2, ~ . + FT.)
m4 <- update(m2, ~ . + Pos)
mtable(m1, m2, m3, m4)
```

The model can only explain 29.1% of the variance in the point per minute. There
might be a feature which we have missed, or not exactly keeping track of as 
a statistics right now. Using per minute stat does not change the r-squared much
from adding both field goals attempted and field goals made, in place of per 
minute stats.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Position, generally, posed no noticable difference for points per minute. 
However, the correlation of field goal percentage and points per minute on 
centers were really small by comparison.

In other positions, field goal percentage generally went with points per minute.


### Were there any interesting or surprising interactions between features?

Field goal percentage by position. Centers had really high field goal percentage
and I think closer shots taken by the centers are the reason for it.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I've tried modelling with linear regression, but r-squared is .29 - there should
be some major feature / indicator that we missed, or could not find from the
current stats.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x=PPM), data = df) +
  ggtitle("Points Per Minute") +
  geom_histogram(binwidth = .01) +
  scale_x_continuous(limits = c(quantile(df$PPM, 0.025, na.rm = TRUE), quantile(df$PPM, 0.975, na.rm = TRUE))) + 
  xlab("Points Per Minute") + 
  ylab("Number of Athletes")
```

### Description One
The distribution of points per minute shows a unimodal data, indicating that the
athletes have normal-like distribution for their expected points per minute.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x=FG., y=PPM), data = df) +
  ggtitle("Points Per Minute by Field Goal Percentage,\n Segmented by Position") +
  scale_x_continuous(limits = c(0.25, 0.6)) +
  scale_y_continuous(limits = c(0.0, 1.0)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm")+
  facet_wrap(~Pos) + 
  ylab("Points Per Minute") +
  xlab("Field Goal Percentage")
```

### Description Two
While points per minute and field goal percentage generally have a positive 
correlation, for centers it is rather neutral, with almost flat correlation.
For the centers, their field goal percentage is bunched up to the right side of 
our x-axis, meaning that they generally have higher field goal percentage, and
their spread is a bit wider, showing no consistent relationship.

### Plot Three
```{r echo=FALSE, Plot_Three}
p1 <- ggplot(aes(x=Pos,y=FG., fill = Pos), data = df) +
  ggtitle("Field Goal Percentage by Position") +
  geom_boxplot() +
  ylab("Field Goal Percentage")+
  scale_fill_brewer(type = 'qual')
p2 <- ggplot(aes(x=FG.), data = df) +
  ggtitle("Field Goal Percentage Density by Position") +
  geom_density(aes(color = Pos)) +
  xlab("Field Goal Percentage")+
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Positions', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
grid.arrange(p1, p2, ncol = 1)
```

### Description Three
Athletes with position Center(C) has the highest median field goal percentage.
A greater proportion of centers have higher field goal percentage. There are 
less outliers in center position, as well.

------

# Reflection

This dataset contains information on all players and their stats from 1950 - 
2017(partial). However, to better reflect modern basketball, I've only used 
2016-2017 as my sample. I started by looking through individual variables, 
making estimates on which variable might be impactful on my main variable, which
was point per minute. 

The trend between field goal percentage and points per minute was found, albeit
not very strong. I was surprised by the fact that the position actually had
a big impact on field goal percentage. Centers were, in general, more efficient
in scoring field goals then others. I hoped I have a bit more data to dive in 
and elaborate on this phenomenon. Average shooting distance, for example, would
be a great feature to have and analyze - I'm quite sure this is what contributed
to the scoring efficiency of the centers.

There are some limitations with this approach: This season may be an abnormality
when compared to other seasons. Also, it is based on the game statistics, which
can be bolstered by the addition of several features. Also, my model was quite 
off, primarily due to the data having great spread in terms of field goal 
percentage and points per minute.